Semantic Linearity in a Lazy Optimising Compiler

### 2024-02-15 icfp.tex
RM edits Sec. 3.5

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
In our system, groups of
recursive let bindings are always assumed to be strongly connected, that is,
all the bindings in a recursive let group are mutually recursive in the sense
that they all (transitively) depend on one another.
%
As before, recursive let bindings bind expressions \emph{lazily}, so they similarly
introduce a $\D$-variable for each binding, and resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-variables or directly, if the let-bindings are unused.
%
However, as discussed in Section
\ref{sec:semantic-linearity-examples:recursive-lets}, we must consider how
recursive uses of a binder in its own definition entails consuming all resources otherwise
required to type the binder's body.
%, i.e. a least upper bound.

Extrapolating to a strongly-connected group of recursive bindings, (mutually)
recursive uses of binders entail consuming all resources
required to type those binders. By definition, those binders in turn
recursively use binders that used them, and thus all the resources otherwise
required to type them.
%
Ultimately, all binders in a strongly-connected group of mutually recursive let
bindings have to be typed with the same linear resources (which are the least
upper bound of resources needed to type the bodies of all binders, accounting
for uses of mutually-recursive binders).
%
% Therefore, we conclude the least upper bound of resources required to type a
% mutually recursive group of let bindings to be the same for any such group that
% is strongly-connected.

The typing rule for recursive groups of bindings leverages our assumption that
all recursive let bindings are strongly connected and the observation
that every binder in a strongly connected group of recursive bindings is typed
with the same linear context. Consequently, all bindings of the recursive group
are introduced as $\D$-vars with the same $\D$ environment -- using any one of
the bindings in a recursive group entails consuming all resources required to
type that same group, which is also why we can use the same linear resources to
type each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
because determining a particular $\D$
to type and annotate all binders is not obvious. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

In practice, determining this typing environment $\D$ amounts to finding a
least upper bound of the resources needed to type each mutually-recursive
binding that (transitively) uses all binders in the recursive group.
%
We propose a naive $O(n^2)$ algorithm for inferring usage environments of recursive
bindings in Section~\ref{sec:discuss:implementation}, orthogonally to the theory
developed in this section.
%
% The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
% environments} used to type each binding.
%
Inference of usage environments for recursive binding groups bears some
resemblance to the inference of principle types for recursive bindings
traditionally achieved through the Hindley–Milner inference
algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to
% develop a better algorithm leveraging existing inference techniques.
% %
% Despite being a seemingly useful observation, 
We leave the exploration of this connection
as future work.

>>>>

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
%
Recursive let bindings also bind expressions \emph{lazily}, so, similarly, we
introduce a $\D$-variable for each binding. Resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-variables, or, directly, if the let-bindings are unused.
%
Additionally, as seen in Section
\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
binder in its own definition entail consuming all resources otherwise required
to type the binder's body.

As for groups of recursive let bindings, we recall that, in our
system, all recursive let groups are strongly-connected -- all bindings in a
recursive let group are mutually recursive in the sense that they all
(transitively) depend on one another.
%
% In light of it, (mutually) recursive uses of binders entail consuming all the
% resources required to type all of the binders.
%
By definition of strongly-connected, any binder in the mutually recursive group
may end up using any binder from the group. Accordingly, occurrences of any
such binder must be typed conservatively, using up the sum of resources needed
to type each of the binders,
%
i.e., all binders in a strongly-connected group of mutually recursive let
bindings have to be typed with the same linear resources (which are the least
upper bound of resources needed to type the bodies of all binders, accounting
for uses of mutually-recursive binders).
%
% Therefore, we conclude the least upper bound of resources required to type a
% mutually recursive group of let bindings to be the same for any such group that
% is strongly-connected.

The typing rule for recursive groups of bindings leverages both the assumption
that all recursive let bindings are strongly connected and the observation that
every binder in a strongly connected group of recursive bindings is typed with
the same linear context. Consequently, all bindings of the recursive group are
introduced as $\D$-vars with the same $\D$ environment -- using any one of the
bindings in a recursive group entails consuming all resources required to type
that same group, which is also why we can use the same linear resources to type
each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
because determining a particular $\D$
to type and annotate all binders is not obvious. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

In practice, determining this typing environment $\D$ amounts to finding a
least upper bound of the resources needed to type each mutually-recursive
binding that (transitively) uses all binders in the recursive group.
%
We propose a naive $O(n^2)$ algorithm for inferring usage environments of recursive
bindings in Section~\ref{sec:discuss:implementation}, orthogonally to the theory
developed in this section.
%
% The algorithm is a $O(n^2)$ traversal over the so-called \emph{naive usage
% environments} used to type each binding.
%
However, we note that inference of usage environments for recursive binding
groups bears some resemblance to the inference of principle types for recursive
bindings traditionally achieved through the Hindley–Milner inference
algorithm~\cite{DBLP:conf/popl/DamasM82}. % , there might be an opportunity to
% develop a better algorithm leveraging existing inference techniques.
% %
% Despite being a seemingly useful observation, 
We leave the exploration of this connection
as future work.
